---
title: "PA1_template"
author: "Tuomas Hjelt"
date: "26.7.2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
```

# Course Project 1

## Load the data and convert date column to proper type 
Loads the needed libraries.
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
```

Reads the csv file to a variable while filtering out missing steps-values and converting date-column to proper type.
```{r}
data_no_na <- read.csv("activity.csv", header = TRUE, sep = ",") %>%
        filter(!is.na(steps)) %>%
        mutate(date = as.Date(as.character(date), format = "%Y-%m-%d"))
```


## What is mean total number of steps taken per day?
Summarizes total steps per date and assigns it to a variable.
```{r}
stepsperday <- data_no_na %>%
      group_by(date) %>%
      summarize(total_steps = sum(steps))
```

Draws a histogram of total number of steps per day.
```{r}
hist(stepsperday$total_steps, main="Histogram of total number of steps per day", 
     xlab="Total number of steps per day")
```

Calculates the mean and median of total steps per day.
```{r}
mean(stepsperday$total_steps)
median(stepsperday$total_steps)
```

## What is the average daily activity pattern?
Summarizes the mean of steps per interval.
```{r}
average_steps <- data_no_na %>%
      group_by(interval) %>%
      summarize(avg_steps = mean(steps))
```

Draws a time series plot of average number of steps per interval.
```{r}
plot(average_steps$interval, average_steps$avg_steps, type = "l", main = "Average number of steps per interval",
      xlab = "Time intervals", ylab = "Average number of steps")
```

Finds the 5-minute interval that contains the maximum number of steps.
```{r}
average_steps %>% 
      filter(avg_steps == max(average_steps$avg_steps)) %>% 
      select(interval)
```

## Imputing missing values
Reads the data with missing values.
```{r}
data <- read.csv("activity.csv", header = TRUE, sep = ",") %>%
  mutate(date = as.Date(as.character(date), format = "%Y-%m-%d"))
```

Counts the total number of missing values.
```{r}
data %>% filter(is.na(steps)) %>% summarize(total_missing_rows = n())
```

### Replace the missing NA values with the average steps in that interval across all the days
I decided to replace the missing NA values with the average steps in that interval across all the days. First I summarized the average steps per interval into a new variable.
```{r}
stepsperinterval <- data_no_na %>%
  group_by(interval) %>%
  summarize(avg_steps = mean(steps))
```

Copies the original dataset and loops through it. When it finds a missing value it gets the average step count according to the right interval from the just created dataset.
```{r}
data_imputed <- data
for (i in 1:nrow(data_imputed))
{
  if(is.na(data_imputed$steps[i]))
  {
    imp_value <- stepsperinterval$avg_steps[which(stepsperinterval$interval == data_imputed$interval[i])]
    data_imputed$steps[i] <- imp_value
  }
}
```

Summarizes the imputed date by calculating the sum of steps per date.
```{r}
stepsperday_imputed <- data_imputed %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps))
```

DraWs a histogram of total number of steps per day with the imputed data.
```{r}
hist(stepsperday_imputed$total_steps, main="Histogram of total number of steps per day (imputed data)", 
     xlab="Total number of steps per day")
```

## Are there differences in activity patterns between weekdays and weekends?
Creates a function to separate weekdays and weekends. (my locale is Finnish so days are also in Finnish)
```{r}
weekday_part <- function(date_val) {
  x <- weekdays(date_val)
  if (x == "lauantai" || x == "sunnuntai") {
    ret_val <- "Weekend"
  }
  else {
    ret_val <- "Weekday"
  }
  ret_val
}
```

Uses sapply to loop through the dates and assigns a new column for the day type. Then summarizes the average steps per interval and day type.
```{r}
data_imputed$day_type <- as.factor(sapply(data_imputed$date, weekday_part))
stepsperday_imputed_wk <- data_imputed %>% group_by(interval, day_type) %>% summarize(avg_steps = mean(steps))
```

Draws two graphs to find differences between weekdays and weekends. 
```{r}
p <- ggplot(stepsperday_imputed_wk, aes(interval,avg_steps)) +
  geom_line(stat = "identity", aes(colour = day_type)) +
  theme_bw() +
  facet_grid(day_type ~ .) +
  labs(x = "Interval", y = "Number of steps") +
  ggtitle("Number of steps per interval per day type")
print(p)
```
There seems to be some variation between them, especially during the mornings where more steps are been taken on weekdays.